---
title: "Analysis on SeedPred dataset"
output:
  html_document:
    df_print: paged
---

<style>
  body {text-align: justify}
</style>

### Passabì Daniele

---

## Assignment

Analyse the dataset *SeedPred* contained in the package `emdbook`, data on seed
predation from Duncan and Duncan (2000) that quantify how many times seeds
of two different species disappeared (presumably taken by seed predators) from
observation stations in Kibale National Park, Uganda. The two species 
(actually the smallest- and largest-seeded species of a set of eight species) are
*Polyscias fulva* (`pol`: seed mass < 0.01 g) and Pseudospondias microcarpa (`psd`:
seed mass 50 g).

The main questions to address are: does the probability of seed removal vary
as a function of distance from the forest edge (10 or 25 m)? Does it depend on
species, possibly as a function of seed mass? On time?

It must be remembered that the numbers of seeds present in each location is
small: 5 initially, then fewer as some are predated. The variable `available` 
indicates how many seeds were available at each time and station to be predated;
of course, the observations with `available=0` do not give any information.
Hence, the first question to address is the distribution of seeds removed; can it
be fitted with the binomial distribution?

The book by Bolker gives many ideas about the analysis, but of course I expect
that you develop your own code.

---

## Analysis

### Libraries

```{r message=FALSE, warning=FALSE}

library(emdbook) # dataset library
library(DT)      # nicer print of dataframes
library(tidyverse)

```

### A glance at the dataset

```{r}

df = SeedPred

# exploration of initial dataframe
datatable(df)

# what are the types of the columns?
glimpse(df)

```

```{r}



```







