---
title: "Analysis on SeedPred dataset"
output:
  html_document:
    df_print: paged
---

<style>
  body {text-align: justify}
</style>

### Passabì Daniele

---

## Assignment

Analyse the dataset *SeedPred* contained in the package `emdbook`, data on seed
predation from Duncan and Duncan (2000) that quantify how many times seeds
of two different species disappeared (presumably taken by seed predators) from
observation stations in Kibale National Park, Uganda. The two species 
(actually the smallest- and largest-seeded species of a set of eight species) are
*Polyscias fulva* (`pol`: seed mass < 0.01 g) and *Pseudospondias microcarpa* (`psd`:
seed mass 50 g).

The main questions to address are: does the probability of seed removal vary
as a function of distance from the forest edge (10 or 25 m)? Does it depend on
species, possibly as a function of seed mass? On time?

It must be remembered that the numbers of seeds present in each location is
small: 5 initially, then fewer as some are predated. The variable `available` 
indicates how many seeds were available at each time and station to be predated;
of course, the observations with `available=0` do not give any information.
Hence, the first question to address is the distribution of seeds removed; can it
be fitted with the binomial distribution?

The book by Bolker gives many ideas about the analysis, but of course I expect
that you develop your own code.

---

## Analysis

### Libraries and setup

```{r message=FALSE, warning=FALSE}

library(emdbook)    # dataset library
library(DT)         # nicer display of dataframes
library(tidyverse)
library(ggplot2)    # plotting graphs
library(hrbrthemes) # theme for plotting
library(cowplot)    # useful for plots side by side

general_col = "#BAD7F2"
secondary_col = "#247BA0"
pol_col = "#F2BAC9"
psd_col = "#F2E2BA"

significant_col = "#70AE6E"
not_significant_col = "#F4743B"

Sys.setlocale("LC_ALL", "English") # we want the plots to be in English

```

### A glance at the dataset
A little exploratory analysis to understand what kind of dataset we are dealing with.

```{r}

df = SeedPred

```


- `station` a factor specifying the station number
- `species` a factor with levels `abz`, `cd`, `cor`, `dio`, `mmu`, `pol`, `psd`, `uva`
- `date` sample date
- `seeds` number of seeds present
- `tcum` cumulative time elapsed
- `tint` time since last sample
- `taken` seeds removed since last sample
- `dist` distance from forest edge (m)
- `available` how many seeds were available at each time and station to be predated

```{r}

# what are the types of the columns?
glimpse(df)

# exploration of initial dataframe
datatable(df, rownames = FALSE)

```

### Cleaning of the dataset
We keep only the information about the two species we are interested in: `pol` and `psd`.

Furthermore, as is made explicit by the text of the delivery, the rows with 
`available=0` do not provide any information, so we remove them (as we do with NAs).

```{r}

df = df[df$species == "pol" | df$species == "psd",] # keep the two species
df = df[df$available != 0 & !is.na(df$available),]  # remove rows where available = 0 or NA
df_clean = df[!is.na(df$seeds),]                    # remove rows where seeds is NA

datatable(df_clean, rownames = FALSE)

```

How many entries for each species and distance?

```{r message=FALSE, warning=FALSE, fig.align = 'center'}

# Find the frequencies of the species
df_freqs = df_clean %>%  
  group_by(species, dist) %>% 
  summarise(n = n())

# Nicer names
df_freqs$species = ifelse(df_freqs$species == "pol", "Polyscias fulva (pol)", "Pseudospondias microcarpa (psd)")
df_freqs$dist = ifelse(df_freqs$dist == 10, "10 meters", "25 meters")

# Plot the results
ggplot(df_freqs, aes(x=species, y=n, fill=dist)) + 
  geom_bar(stat = "identity", width = 0.6, position = position_dodge()) +
  theme_ipsum() +
  labs(title = "Occurrences of each species",
       x = "Species", y = "Occurrences") + 
  scale_fill_manual("legend", values = c("10 meters" = general_col, "25 meters" = secondary_col)) + 
  theme(legend.title = element_blank()) +
  geom_text(aes(label=n), position=position_dodge(width=0.6), vjust=1.5)

```

A simple plot of the data, for one station.

```{r message=FALSE, warning=FALSE, fig.align = 'center'}

example_station = 19
example_distance = 25

df_stat_65 = df_clean[df_clean$station == example_station & df_clean$dist == example_distance,]

df_stat_65 %>%
  ggplot( aes(x=date, y=seeds)) +
    geom_line(color="grey") +
    geom_point(shape=21, color="black", fill=pol_col, size=6) +
    theme_ipsum() +
    labs(title = "Seeds predation through time",
         subtitle = "Station: 19\nSpecies: Polyscias fulva (pol)\nDistance: 25m",
         x = "Date", y = "Seeds")

```

Now we generalize this plot.

```{r message=FALSE, warning=FALSE}

# CASE 1: pol, 10
df_pol_10 = df_clean[df_clean$species == "pol" & df_clean$dist == 10,]

# count how many measuraments there are for each date
df_temp_1 = df_pol_10 %>%
  count(date)

# count the total number of seeds for each date
df_temp_2 = df_pol_10 %>%
  group_by(date) %>%
  summarise(total_seeds = sum(seeds))

# merge the information to obtain mean seeds available
df_temp_2$n = df_temp_1$n
df_temp_2$mean_seeds = df_temp_2$total_seeds / df_temp_2$n

df_pol_10 = df_temp_2

###

# CASE 2: pol, 25
df_pol_25 = df_clean[df_clean$species == "pol" & df_clean$dist == 25,]

# count how many measuraments there are for each date
df_temp_1 = df_pol_25 %>%
  count(date)

# count the total number of seeds for each date
df_temp_2 = df_pol_25 %>%
  group_by(date) %>%
  summarise(total_seeds = sum(seeds))

# merge the information to obtain mean seeds available
df_temp_2$n = df_temp_1$n
df_temp_2$mean_seeds = df_temp_2$total_seeds / df_temp_2$n

df_pol_25 = df_temp_2

###

# CASE 3: psd, 10
df_psd_10 = df_clean[df_clean$species == "psd" & df_clean$dist == 10,]

# count how many measuraments there are for each date
df_temp_1 = df_psd_10 %>%
  count(date)

# count the total number of seeds for each date
df_temp_2 = df_psd_10 %>%
  group_by(date) %>%
  summarise(total_seeds = sum(seeds))

# merge the information to obtain mean seeds available
df_temp_2$n = df_temp_1$n
df_temp_2$mean_seeds = df_temp_2$total_seeds / df_temp_2$n

df_psd_10 = df_temp_2

###

# CASE 4: psd, 25
df_psd_25 = df_clean[df_clean$species == "psd" & df_clean$dist == 25,]

# count how many measuraments there are for each date
df_temp_1 = df_psd_25 %>%
  count(date)

# count the total number of seeds for each date
df_temp_2 = df_psd_25 %>%
  group_by(date) %>%
  summarise(total_seeds = sum(seeds))

# merge the information to obtain mean seeds available
df_temp_2$n = df_temp_1$n
df_temp_2$mean_seeds = df_temp_2$total_seeds / df_temp_2$n

df_psd_25 = df_temp_2

```


```{r message=FALSE, warning=FALSE, fig.align = 'center'}

# Now we plot

pol1 = df_pol_10 %>%
  ggplot( aes(x=date, y=mean_seeds)) +
    geom_line(color="grey") +
    geom_point(shape=21, color="black", fill=pol_col, size=2.5) +
    ylim(0, 5) +
    theme_ipsum() +
    labs(title = "Mean seeds available through time",
         subtitle = "Species: Polyscias fulva\nDistance: 10m",
         x = "Date", y = "Seeds")

pol2 = df_pol_25 %>%
  ggplot( aes(x=date, y=mean_seeds)) +
    geom_line(color="grey") +
    geom_point(shape=21, color="black", fill=pol_col, size=2.5) +
    ylim(0, 5) +
    theme_ipsum() +
    labs(title = "",
         subtitle = "Species: Polyscias fulva\nDistance: 25m",
         x = "Date", y = "Seeds")

psd1 = df_psd_10 %>%
  ggplot( aes(x=date, y=mean_seeds)) +
    geom_line(color="grey") +
    geom_point(shape=21, color="black", fill=psd_col, size=2.5) +
    ylim(0, 5) +
    theme_ipsum() +
    labs(title = "Mean seeds available through time",
         subtitle = "Species: Pseudospondias microcarpa\nDistance: 10m",
         x = "Date", y = "Seeds")

psd2 = df_psd_25 %>%
  ggplot( aes(x=date, y=mean_seeds)) +
    geom_line(color="grey") +
    geom_point(shape=21, color="black", fill=psd_col, size=2.5) +
    ylim(0, 5) +
    theme_ipsum() +
    labs(title = "",
         subtitle = "Species: Pseudospondias microcarpa\nDistance: 25m",
         x = "Date", y = "Seeds")

plot_grid(pol1, pol2, labels = "", ncol = 2)
plot_grid(psd1, psd2, labels = "", ncol = 2)

```

From these simple graphs we can see that seed species seems to strongly influence the average number of available seeds. Distance to station, on the other hand, seems to be relevant only for the Pseudospondias microcarpa seed.


### Can the *distribution of seeds removed* be fitted with the *binomial distribution*?

We can interpret in 2 ways our data.

**Case A**

For each `date` *d*, the `available` seeds can either be predated or not 
(we do not consider each individual seed).

So, we create a new column, `seeds_rem`, which is a boolean column where:

- `TRUE` indicates that any number of seeds were predated (based on `taken` column)
- `FALSE` indicates that no seeds were predated

```{r}

df_clean$seeds_rem = ifelse(df_clean$taken == 0, FALSE, TRUE)

datatable(df_clean, rownames = FALSE)
summary(df_clean$seeds_rem)

```

```{r}

# based on summary we know that
successes = 97
n = 890 + 97
emp_prob = successes/n

message("Number of trials: ", n, "\nProbability of seeds removal: ", round(emp_prob,4))

```

```{r message=FALSE, warning=FALSE, fig.align = 'center'}

x1 = 0:n
df = data.frame(x = x1, y = dbinom(x1, n, emp_prob))

plot_full = ggplot(df, aes(x = x, y = y)) + 
  geom_bar(stat = "identity", col = general_col, fill = general_col) +
  xlab("Number of successes") + 
  ylab("Density") +
  ylim(0,0.05) +
  labs(title = "Binomial distribution",
       subtitle = "n=987\np=0.0983") + 
  theme_ipsum() 

plot_zoom = ggplot(df, aes(x = x, y = y)) + 
  geom_bar(stat = "identity", col = general_col, fill = general_col) +
  xlab("Number of successes") + 
  xlim(50,150) +
  ylab("Density") +
  ylim(0,0.05) +
  labs(title = "Binomial distribution",
       subtitle = "n=987\np=0.0983") + 
  theme_ipsum() 

plot_full
plot_zoom

```

**Case B**

We consider each seed individually, so the focus lies on the columns `taken` and `available`.

```{r}

successes = sum(df_clean$taken)
n = sum(df_clean$available)
emp_prob = successes/n

message("Number of trials: ", n, "\nNumber of successes: ", successes, "\nProbability of seeds removal: ", round(emp_prob,4))

```

```{r message=FALSE, warning=FALSE, fig.align = 'center'}

x1 = 0:n
df = data.frame(x = x1, y = dbinom(x1, n, emp_prob))

plot_full = ggplot(df, aes(x = x, y = y)) + 
  geom_bar(stat = "identity", col = general_col, fill = general_col) +
  xlab("Number of successes") + 
  ylab("Density") +
  ylim(0,0.03) +
  labs(title = "Binomial distribution",
       subtitle = "Focus on each individual seed\n\nn=3620\np=0.0685") + 
  theme_ipsum() 

plot_zoom = ggplot(df, aes(x = x, y = y)) + 
  geom_bar(stat = "identity", col = general_col, fill = general_col) +
  xlab("Number of successes") + 
  xlim(150,350) +
  ylab("Density") +
  ylim(0,0.03) +
  labs(title = "Binomial distribution",
       subtitle = "Focus on each individual seed\n\nn=3620\np=0.0685") + 
  theme_ipsum() 

plot_full
plot_zoom

```



### Linear models
#### Does the probability of seed removal vary as a function of distance from the forest edge (10 or 25 m)? 
#### Does it depend on species, possibly as a function of seed mass? On time?

```{r}

# Create the column with the probability of seed removal [p_seed_rem]

df_clean$p_seed_rem = round(df_clean$taken / df_clean$available, 4)

datatable(df_clean, rownames = FALSE)

```

---

#### Distance
Is it significant?

```{r}

model_dist = lm(p_seed_rem~dist, data=df_clean)

# When we use a *qualitative variable* the best way to look at the data is *anova*
print(anova(model_dist))

```

<p style="color:#F4743B">**`## dist        1  0.146 0.145519  2.7861 0.0954 .`**</p>


```{r}

# Other useful info can be obtained through summary
summary(model_dist)

```

<p style="color:#F4743B">**`## dist25      0.024352   0.014589   1.669   0.0954 .`**</p>

From the result we can see how the **distance** is not that significant. 
There is *no substantial difference* between the reference value (10 meters) and 25m.

---

#### Species
Are they significant?

```{r}

model_species = lm(p_seed_rem~species, data=df_clean)

print(anova(model_species))

```

<p style="color:#70AE6E">**`## species     1  2.874 2.87353  58.097 5.855e-14 ***`**</p>


```{r}

summary(model_species)

```

<p style="color:#70AE6E">**`## speciespsd  -0.12632    0.01657  -7.622 5.86e-14 ***`**</p>


The p-value is extremely low, so we can say with a high degree of confidence 
that the variable **species** is relevant. This result confirms what we hypothesized in the exploratory analysis.

We can get more useful information through the use of the *residuals plot*.

```{r fig.align = 'center'}

par(mfrow = c(2,2))
plot(model_species)

```

It is not necessary to analyze in depth all the graphs to understand that the model 
is not extremely accurate. 

In the first and in the third graph we immediately notice how the data is not homogeneously distributed.
In the second one it is obvious that a normal distribution is not followed. 
The fourth does not detect the presence of outliers.

---

#### Time
Is it significant?
We use the variable `tcum`, the cumulative time elapsed.

```{r}

model_time = lm(p_seed_rem~tcum, data=df_clean)
summary(model_time)

```

<p style="color:#70AE6E">**`## tcum        -8.581e-04  9.175e-05  -9.352   <2e-16 ***`**</p>

In this case, as in the previous one, we get a very low p-value and we can say that the variable `tcum` is significant.

```{r fig.align = 'center'}

par(mfrow = c(2,2))
plot(model_time)

```

The graphs seem to be a little bit nicer, but we can achieve better results.

---

#### Date

It is reasonable to use the date, but instead of focusing on the single day, 
better results should be achieved using months. So, we do a little bit of data manipulation and we add the column `month` to our dataframe.

```{r}

df_clean$month = as.factor(months(df_clean$date))
datatable(df_clean, rownames = FALSE)

```


```{r}

model_month = lm(p_seed_rem~month, data=df_clean)

print(anova(model_month))

```

<p style="color:#70AE6E">**`## month       8  5.956 0.74451  15.955 < 2.2e-16 ***`**</p>

According to the Analysis of Variance Table, the variable `month` is significant. 
Let's investigate further.

```{r}

summary(model_month)

```

Our `ANOVA` results seems to be confirmed. There is a difference between most of the months. We can confirm this even further, using `TukeyHSD`.

```{r}

# TukeyHSD computes the Honest significance differences and works on the output of aov
TukeyHSD(aov(model_month))

```

As always, we plot the residuals.

```{r fig.align = 'center'}

par(mfrow = c(2,2))
plot(model_month)

```

---

#### Summary of the models
We use AIC (Akaike Information Criterion) and BIC (Bayes Information Criterion) to evaluate the models we have created so far.

```{r}

# create a simple df
col_names = c("model", "AIC", "BIC")

models = c("distance", "species", "cumulative time", "month")
AIC = c(AIC(model_dist), AIC(model_species), AIC(model_time), AIC(model_month))
BIC = c(BIC(model_dist), BIC(model_species), BIC(model_time), BIC(model_month))

eval_df = data.frame(models, AIC, BIC)

datatable(eval_df, rownames = FALSE, class = 'cell-border stripe')

```


---

#### Experiments with Models 

We would like to try to fit a model using all variables 
(chosen from those that are reasonable to use).

```{r}

model_exp_1 = lm(p_seed_rem~dist+species+month+tcum+available, data=df_clean)
summary(model_exp_1)

```

<p style="color:#F4743B">**`## dist25          0.0137818  0.0139867   0.985  0.32470`**</p>
<p style="color:#70AE6E">**`## speciespsd     -0.0957068  0.0200467  -4.774 2.08e-06 ***`**</p>
<p style="color:#70AE6E">**`## monthJune      -0.1402674  0.0583903  -2.402  0.01648 *`**</p>
<p style="color:#70AE6E">**`## monthMay       -0.1042007  0.0358869  -2.904  0.00377 **`**</p>
<p style="color:#F4743B">**`## tcum           -0.0002838  0.0008063  -0.352  0.72499`**</p>
<p style="color:#F4743B">**`## available       0.0037999  0.0063611   0.597  0.55040`**</p>

Things are a little bit different with this model. What seems to be relavant are
`species` and `month.`

We remove the distance (`dist`), cumulative time (`tcum`) and `available`.


```{r}

model_exp_2 = lm(p_seed_rem~species+month, data=df_clean)
summary(model_exp_2)

```

Instead of doing it manually, let's try to use `step`. This function takes away one variable from the model each time, choosing the one that decreases most AIC. It stops when taking away variables would only increase AIC.

```{r}

step(model_exp_1) 

```

We obtain the same results as what we did manually (`model_exp_2`).

Let's plot the residuals and evaluate the model.

```{r fig.align = 'center'}

par(mfrow = c(2,2))
plot(model_exp_2)

```

```{r echo=FALSE}

message("Evaluation of Model with SPECIES+MONTH\nAIC: ", AIC(model_exp_2), "\nBIC: ", BIC(model_exp_2))

```

The residuals are not perfect, but according to AIC and BIC this is the best model so far.
























