---
title: "Analysis on SeedPred dataset"
output:
  html_document:
    df_print: paged
---

<style>
  body {text-align: justify}
</style>

### Passabì Daniele

---

## Assignment

Analyse the dataset *SeedPred* contained in the package `emdbook`, data on seed
predation from Duncan and Duncan (2000) that quantify how many times seeds
of two different species disappeared (presumably taken by seed predators) from
observation stations in Kibale National Park, Uganda. The two species 
(actually the smallest- and largest-seeded species of a set of eight species) are
*Polyscias fulva* (`pol`: seed mass < 0.01 g) and *Pseudospondias microcarpa* (`psd`:
seed mass 50 g).

The main questions to address are: does the probability of seed removal vary
as a function of distance from the forest edge (10 or 25 m)? Does it depend on
species, possibly as a function of seed mass? On time?

It must be remembered that the numbers of seeds present in each location is
small: 5 initially, then fewer as some are predated. The variable `available` 
indicates how many seeds were available at each time and station to be predated;
of course, the observations with `available=0` do not give any information.
Hence, the first question to address is the distribution of seeds removed; can it
be fitted with the binomial distribution?

The book by Bolker gives many ideas about the analysis, but of course I expect
that you develop your own code.

---

## Analysis

### Libraries and setup

```{r message=FALSE, warning=FALSE}

library(emdbook)    # dataset library
library(DT)         # nicer display of dataframes
library(tidyverse)
library(ggplot2)    # plotting graphs
library(hrbrthemes) # theme for plotting
library(cowplot)    # useful for plots side by side

general_col = "#BAD7F2"
pol_col = "#F2BAC9"
psd_col = "#F2E2BA"

Sys.setlocale("LC_ALL", "English") # we want the plots to be in English

```

### A glance at the dataset
A little exploratory analysis to understand what kind of dataset we are dealing with.

```{r}

df = SeedPred

```


- `station` a factor specifying the station number
- `species` a factor with levels `abz`, `cd`, `cor`, `dio`, `mmu`, `pol`, `psd`, `uva`
- `date` sample date
- `seeds` number of seeds present
- `tcum` cumulative time elapsed
- `tint` time since last sample
- `taken` seeds removed since last sample
- `dist` distance from forest edge (m)
- `available` how many seeds were available at each time and station to be predated

```{r}

# what are the types of the columns?
glimpse(df)

# exploration of initial dataframe
datatable(df)

```

### Cleaning of the dataset
We keep only the information about the two species we are interested in: `pol` and `psd`.

Furthermore, as is made explicit by the text of the delivery, the rows with 
`available=0` do not provide any information, so we remove them (as we do with NAs).

```{r}

df = df[df$species == "pol" | df$species == "psd",] # keep the two species
df = df[df$available != 0 & !is.na(df$available),]  # remove rows where available = 0 or NA
df_clean = df[!is.na(df$seeds),]                    # remove rows where seeds is NA

datatable(df_clean)

```

A simple plot of the data, for one station.

```{r message=FALSE, warning=FALSE, fig.align = 'center'}

example_station = 19
example_distance = 25

df_stat_65 = df_clean[df_clean$station == example_station & df_clean$dist == example_distance,]

df_stat_65 %>%
  ggplot( aes(x=date, y=seeds)) +
    geom_line(color="grey") +
    geom_point(shape=21, color="black", fill=pol_col, size=6) +
    theme_ipsum() +
    labs(title = "Seeds predation through time",
         subtitle = "Station: 19\nSpecies: Polyscias fulva (pol)\nDistance: 25m",
         x = "Date", y = "Seeds")

```

Now we generalize this plot.

```{r message=FALSE, warning=FALSE}

# CASE 1: pol, 10
df_pol_10 = df_clean[df_clean$species == "pol" & df_clean$dist == 10,]

# count how many measuraments there are for each months
df_temp_1 = df_pol_10 %>%
  count(date)

# count the total number of seeds for each month
df_temp_2 = df_pol_10 %>%
  group_by(date) %>%
  summarise(total_seeds = sum(seeds))

df_temp_2$n = df_temp_1$n
df_temp_2$mean_seeds = df_temp_2$total_seeds / df_temp_2$n

df_pol_10 = df_temp_2

###

# CASE 2: pol, 25
df_pol_25 = df_clean[df_clean$species == "pol" & df_clean$dist == 25,]

# count how many measuraments there are for each months
df_temp_1 = df_pol_25 %>%
  count(date)

# count the total number of seeds for each month
df_temp_2 = df_pol_25 %>%
  group_by(date) %>%
  summarise(total_seeds = sum(seeds))

df_temp_2$n = df_temp_1$n
df_temp_2$mean_seeds = df_temp_2$total_seeds / df_temp_2$n

df_pol_25 = df_temp_2

###

# CASE 3: psd, 10
df_psd_10 = df_clean[df_clean$species == "psd" & df_clean$dist == 10,]

# count how many measuraments there are for each months
df_temp_1 = df_psd_10 %>%
  count(date)

# count the total number of seeds for each month
df_temp_2 = df_psd_10 %>%
  group_by(date) %>%
  summarise(total_seeds = sum(seeds))

df_temp_2$n = df_temp_1$n
df_temp_2$mean_seeds = df_temp_2$total_seeds / df_temp_2$n

df_psd_10 = df_temp_2

###

# CASE 4: pol, 25
df_psd_25 = df_clean[df_clean$species == "psd" & df_clean$dist == 25,]

# count how many measuraments there are for each months
df_temp_1 = df_psd_25 %>%
  count(date)

# count the total number of seeds for each month
df_temp_2 = df_psd_25 %>%
  group_by(date) %>%
  summarise(total_seeds = sum(seeds))

df_temp_2$n = df_temp_1$n
df_temp_2$mean_seeds = df_temp_2$total_seeds / df_temp_2$n

df_psd_25 = df_temp_2

```


```{r message=FALSE, warning=FALSE, fig.align = 'center'}

# Now we plot

pol1 = df_pol_10 %>%
  ggplot( aes(x=date, y=mean_seeds)) +
    geom_line(color="grey") +
    geom_point(shape=21, color="black", fill=pol_col, size=2.5) +
    ylim(0, 5) +
    theme_ipsum() +
    labs(title = "Mean seeds available through time",
         subtitle = "Species: Polyscias fulva\nDistance: 10m",
         x = "Date", y = "Seeds")

pol2 = df_pol_25 %>%
  ggplot( aes(x=date, y=mean_seeds)) +
    geom_line(color="grey") +
    geom_point(shape=21, color="black", fill=pol_col, size=2.5) +
    ylim(0, 5) +
    theme_ipsum() +
    labs(title = "",
         subtitle = "Species: Polyscias fulva\nDistance: 25m",
         x = "Date", y = "Seeds")

psd1 = df_psd_10 %>%
  ggplot( aes(x=date, y=mean_seeds)) +
    geom_line(color="grey") +
    geom_point(shape=21, color="black", fill=psd_col, size=2.5) +
    ylim(0, 5) +
    theme_ipsum() +
    labs(title = "Mean seeds available through time",
         subtitle = "Species: Pseudospondias microcarpa\nDistance: 10m",
         x = "Date", y = "Seeds")

psd2 = df_psd_25 %>%
  ggplot( aes(x=date, y=mean_seeds)) +
    geom_line(color="grey") +
    geom_point(shape=21, color="black", fill=psd_col, size=2.5) +
    ylim(0, 5) +
    theme_ipsum() +
    labs(title = "",
         subtitle = "Species: Pseudospondias microcarpa\nDistance: 25m",
         x = "Date", y = "Seeds")

plot_grid(pol1, pol2, labels = "", ncol = 2)
plot_grid(psd1, psd2, labels = "", ncol = 2)

```

### Can the *distribution of seeds removed* be fitted with the *binomial distribution*?

We can interpret in 2 ways our data.

**Case A**

For each `date` *d*, the `available` seeds can either be predated or not 
(we do not consider each individual seed).

So, we create a new column, `seeds_rem`, which is a boolean column where:

- `TRUE` indicates that any number of seeds were predated (based on `taken` column)
- `FALSE` indicates that no seeds were predated

```{r}

df_clean$seeds_rem = ifelse(df_clean$taken == 0, FALSE, TRUE)

datatable(df_clean)
summary(df_clean$seeds_rem)

```

```{r}

# based on summary we know that
successes = 97
n = 890 + 97
emp_prob = successes/n

message("Number of trials: ", n, "\nProbability of seeds removal: ", round(emp_prob,4))

```

```{r message=FALSE, warning=FALSE, fig.align = 'center'}

x1 = 0:n
df = data.frame(x = x1, y = dbinom(x1, n, emp_prob))

plot_full = ggplot(df, aes(x = x, y = y)) + 
  geom_bar(stat = "identity", col = general_col, fill = general_col) +
  xlab("Number of successes") + 
  ylab("Density") +
  ylim(0,0.05) +
  labs(title = "Binomial distribution",
       subtitle = "n=987\np=0.0983") + 
  theme_ipsum() 

plot_zoom = ggplot(df, aes(x = x, y = y)) + 
  geom_bar(stat = "identity", col = general_col, fill = general_col) +
  xlab("Number of successes") + 
  xlim(50,150) +
  ylab("Density") +
  ylim(0,0.05) +
  labs(title = "Binomial distribution",
       subtitle = "n=987\np=0.0983") + 
  theme_ipsum() 

plot_full
plot_zoom

```

**Case B**

We consider each seed individually, so the focus lies on the columns `taken` and `available`.

```{r}

successes = sum(df_clean$taken)
n = sum(df_clean$available)
emp_prob = successes/n

message("Number of trials: ", n, "\nNumber of successes: ", successes, "\nProbability of seeds removal: ", round(emp_prob,4))

```

```{r message=FALSE, warning=FALSE, fig.align = 'center'}

x1 = 0:n
df = data.frame(x = x1, y = dbinom(x1, n, emp_prob))

plot_full = ggplot(df, aes(x = x, y = y)) + 
  geom_bar(stat = "identity", col = general_col, fill = general_col) +
  xlab("Number of successes") + 
  ylab("Density") +
  ylim(0,0.03) +
  labs(title = "Binomial distribution",
       subtitle = "Focus on each individual seed\n\nn=3620\np=0.0685") + 
  theme_ipsum() 

plot_zoom = ggplot(df, aes(x = x, y = y)) + 
  geom_bar(stat = "identity", col = general_col, fill = general_col) +
  xlab("Number of successes") + 
  xlim(150,350) +
  ylab("Density") +
  ylim(0,0.03) +
  labs(title = "Binomial distribution",
       subtitle = "Focus on each individual seed\n\nn=3620\np=0.0685") + 
  theme_ipsum() 

plot_full
plot_zoom

```



### Linear models
#### Does the probability of seed removal vary as a function of distance from the forest edge (10 or 25 m)? 
#### Does it depend on species, possibly as a function of seed mass? On time?

```{r}

# Create the column with the probability of seed removal [p_seed_rem]

df_clean$p_seed_rem = df_clean$taken / df_clean$available

datatable(df_clean)

```

---

#### Distance

```{r}

# DISTANCE - is it significant?

model_dist = lm(p_seed_rem~dist, data=df_clean)
summary(model_dist)

```

From the result we can see how the **distance** is not that significant. 
There is *no substantial difference* between the reference value (10 meters) and 25m.

---

#### Species

```{r}

# SPECIES - is it significant?

model_species = lm(p_seed_rem~species, data=df_clean)
summary(model_species)

```

The p-value is extremely low, so we can say with a high degree of confidence 
that the variable **species** is relevant.

This result confirms what we hypothesized in the exploratory analysis.

We can get more useful information through the use of the *residuals plot*.

```{r}

plot(model_species)

```

It is not necessary to analyze the graphs in detail to understand that the model 
is not extremely accurate. 

In the first and in the third graph we immediately notice how the data is not homogeneous.
In the second one it is obvious that a normal distribution is not followed. 
The fourth does not detect the presence of outliers.

---

#### Time
Is it significant?
We use the variable tcum.

```{r}

model_time = lm(p_seed_rem~tcum, data=df_clean)
summary(model_time)
plot(model_time)

```

The graphs seem to be a little bit nicer, but we can achieve better results. 

---

#### Experiments with Models 

```{r}

# Let's try to fit a model with all the variables that make sense

model_1 = lm(p_seed_rem~dist+species+tcum, data=df_clean)
summary(model_1)

```

As before, the distance is not significant, so we remove it.

```{r}

model_2 = lm(p_seed_rem~species+tcum, data=df_clean)
summary(model_2)

```




























